Verification of Stencil Computations with Spatial Specifications

### 2016-07-07 paper.tex
last edits

From \textsf{schematic} we now give a
formal characterisation of stencil computations for our purposes:
%%
\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all $b \in
  \bar{b}$ and all subscripts $b(\bar{e}') \in e_r . \, $
  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
\label{def:stencil}
\end{definition}
%%

@@

The second equation considers the case where the specification is
an \term{atMost} upper-bound approximation (labelled with \textsf{up}
as shorthand for \textsf{upper} here). 
An upper bound specification means that for every scheme in the code
model, modelling a source-level index, $\forall u \in M_c$ there exists
a generalising scheme in the spec model $\exists v \in M$, \ie{}, $u
\preceq_n v$. That is, every source-level index is generalised by the model.

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in
Section~\ref{subsec:inference} (see top-left):
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

@@

\subsection{Software Corpus}

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\end{tabular}
\caption{Summary of software packages used for
  evaluation\label{tab:corpus}}
\vspace{-1em}
\end{table}
\begin{table}
\vspace{-0.5em}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth   
\end{tabular}
\caption{Stencil specification assignment rates\label{tab:tickAssign}}
\vspace{-1em}
\end{table}

@@

\begin{figure}[t]
\begin{minted}[breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
!=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
\end{minted}
\caption{A complex specification inferred from
  \textbf{Unified Model}\label{fig:smagorinsky}}
\vspace{-1em}
\end{figure}

@@

There are 48 possible errors of this type of which only 6 were not 
detected by the checker. The reason is that the inferred
specification describes a family of access patterns. Some offset errors
produce a stencil computation that belongs to the same family and goes
undetected.
One of the reasons this specification can catch many offset errors is that it is
a \texttt{readOnce} specification. This prevents offset errors leading to
duplicate access patterns be caught even if the resulting set of accesses are
part of the same family.
% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
% 42 cases were caught
% 6 cases were not caught
\section{Related work}
\label{sec:discussion}

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that the high-level abstract specifications can
be automatically translated into high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is a post-condition generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsc{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j);  a(i, j) = q + t
        t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and not to
that on line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).

@@

\section{Conclusion and future work}

There is an increasing awareness
of the need for verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lightweight specification for numerical code.
Compared to other specification approaches (discussed above), our language is novel in that it does
not aim to reify indexing in the specification, but instead
provides abstract spatial descriptions which capture a large number
of common patterns. Our tool is useful for both detecting
errors and for specifying program behaviour to future-proof code and ease
 maintenance.

@@

We evaluated our approach by implementing the stencil inference and
checking procedure on top of CamFort, an open-source analysis
tool for Fortran
programs and running the inference process over a corpus of 1.1
million lines of Fortran code.  Our motivation for this work was to
provide an intermediate specification language to allow scientists to
verify numerical computations.  Our evaluation showed that relatively
simple stencil computations are the norm, for which we can specify a
large proportion, comprising mainly reflexive region constants which
are less likely to give rise to programming errors.  However, around
10\% of stencils covered more complicated program structures for which
indexing errors are much more plausible. In the more detailed cases,
with larger specifications, the importance and impact of such
specifications grows.  We showed two examples of more complicated
stencils, for which CamFort was able to detect a significant majority
of possible indexing errors in our case studies.

>>>>

From \textsf{schematic} we now give a
formal characterisation of stencil computations for our purposes:
%%
\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and %for all $b \in  \bar{b}$ and
  for all subscripts $b(\bar{e}') \in e_r . \, $
  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
\label{def:stencil}
\end{definition}
%%

@@

The second equation considers the case where the specification is
an \term{atMost} upper-bound approximation (labelled with \textsf{up}
as shorthand for \textsf{upper} here). 
An upper bound specification means that for every scheme in the code
model, modelling a source-level index, $\forall u \in M_c$ there exists
a generalising scheme in the spec model $\exists v \in M_s$, \ie{}, $u
\preceq_n v$. That is, every source-level index is generalised by the model.

@@

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in
Section~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

@@

%\subsection{Software Corpus}
\paragraph{Software Corpus}

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\end{tabular}
\caption{Summary of software packages used for
  evaluation\label{tab:corpus}}
\vspace{-1em}
\end{table}
\begin{table}
\vspace{-1em}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth   
\end{tabular}
\caption{Stencil specification assignment rates\label{tab:tickAssign}}
\vspace{-1em}
\end{table}

@@

\begin{figure}[t]\begin{minted}[fontsize=\scriptsize,breakindent=0em,linenos=false,xleftmargin=0em,breakafter=)]{fortran}
!=stencil readOnce,(forward(depth=1,dim=3,irreflexive))*(backward(depth=1,dim=1))*(backward(depth=1,dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1))*(reflexive(dim=2))+(forward(depth=1,dim=3))*(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2,irreflexive))+(forward(depth=1,dim=3))*(backward(depth=1,dim=2))*(reflexive(dim=1))+(backward(depth=1,dim=1))*(backward(depth=1,dim=2,irreflexive))*(reflexive(dim=3))+(backward(depth=1,dim=1,irreflexive))*(backward(depth=1,dim=2))*(reflexive(dim=3))::x
\end{minted}
\caption{A complex specification inferred from
  \textbf{Unified Model}\label{fig:smagorinsky}}
\vspace{-1em}
\end{figure}

@@

There are 48 possible errors of this type of which only 6 were not 
detected by the checker. The reason is that the inferred
specification describes a family of access patterns. Some offset errors
produce a stencil computation that belongs to the same family and goes
undetected.
One of the reasons this specification can catch many offset errors is that it is
a \texttt{readOnce} specification. This prevents offset errors leading to
duplicate access patterns be caught even if the resulting set of accesses are
part of the same family.
% /um/trunk/src/atmosphere/diffusion_and_filtering/turb_smagorinsky.F90
% Tried introducing indexing errors replacing one of the indices of cx_rho_instance with -1,0,+1
% 42 cases were caught
% 6 cases were not caught
\section{Related work and conclusions}
\label{sec:discussion}

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that the high-level abstract specifications can
be automatically translated into high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is a post-condition generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsc{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j);  a(i, j) = q + t; t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 4 maps only to the definition on line 2, and not to
that on line 4. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).

@@

\paragraph{Conclusion and future work}

There is an increasing awareness
of the need for verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lightweight specification for numerical code.
Compared to other specification approaches (discussed above), our language is novel in that it does
not aim to reify indexing in the specification, but instead
provides abstract spatial descriptions which capture a large number
of common patterns. %Our tool is useful for both detecting
%errors and for specifying program behaviour to future-proof code and ease
% maintenance.

@@

We evaluated our approach on top of CamFort, an open-source analysis
tool for Fortran
programs, running the inference process over a corpus of 1.1
million lines of Fortran code. 
 Our motivation for this work was to
provide an intermediate specification language to allow scientists to
verify numerical computations.  Our evaluation showed that relatively
simple stencil computations are the norm comprising mainly reflexive region constants which
are less likely to give rise to programming errors.  However, around
10\% of stencils covered more complicated program structures for which
indexing errors are much more plausible. In the more detailed cases,
with larger specifications, the importance and impact of such
specifications grows.  We showed two examples of more complicated
stencils, for which CamFort was able to detect a significant majority
of possible indexing errors in our case studies.
